---
title: "From Data to Impact: Interactive Analysis of HIV and Child Mortality in the EAC(2000-2023)"
author: "Joseph Kimani Mbugua"
date: "2025-04-2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(readxl)
library(dplyr)
library(stringr)
library(ggplot2)
library(plotly)
library(scales)
library(sf)
library(lme4)
library(corrplot)
```

## Data Cleaning

### HIV Data Cleaning
load and clean the WHO HIV dataset (2000-2023).

```{r load_hiv_data}
hiv_data <- read_excel("HIV data 2000-2023.xlsx")

# Function to extract central HIV estimate
extract_estimate <- function(value) {
  if (value == "No data") {
    return(NA)
  } else if (str_detect(value, "<")) {
    num <- str_replace(value, "<", "")
    return(as.numeric(num))
  } else {
    part <- str_split(value, "\\[")[[1]][1]
    part <- str_trim(part)
    part <- str_replace_all(part, " ", "")
    return(as.numeric(part))
  }
}

# Apply function and rename columns
hiv_data <- hiv_data %>%
  mutate(HIV_Estimate = sapply(Value, extract_estimate)) %>%
  rename(Country = Location, Year = Period) %>%
  select(Country, Year, HIV_Estimate, ParentLocationCode)
head(hiv_data)
```

### Multidimensional Poverty Data Cleaning
load and clean the World Bank multidimensional poverty dataset.

```{r load_poverty_data}
poverty_data <- read_excel("multidimensional_poverty.xlsx")

# Select relevant columns and rename for clarity
poverty_data <- poverty_data %>%
  select(Economy, `Survey year`, `Multidimensional poverty headcount ratio (%)`, 
         `Monetary (%)`, `Educational attainment (%)`, `Educational enrollment (%)`, 
         `Electricity (%)`, `Sanitation (%)`, `Drinking water (%)`) %>%
  rename(Country = Economy, Year = `Survey year`, MPI = `Multidimensional poverty headcount ratio (%)`, 
         Monetary = `Monetary (%)`, Education_Attainment = `Educational attainment (%)`, 
         Education_Enrollment = `Educational enrollment (%)`, Electricity = `Electricity (%)`, 
         Sanitation = `Sanitation (%)`, Drinking_Water = `Drinking water (%)`)

# Convert percentages to proportions
poverty_data <- poverty_data %>%
  mutate(across(MPI:Drinking_Water, ~ as.numeric(as.character(.)) / 100))
head(poverty_data)
```

### Child Mortality Data Cleaning
load and clean the UN IGME mortality datasets and integrate `dataset_datascience.csv` for EAC countries.

```{r load_mortality_data, include=FALSE}
neonatal_data <- read.csv("neonatal_mortality_rate.csv")
under_five_data <- read.csv("under_five mortality rate.csv")
datascience_data <- read.csv("dataset_datascience.csv")

eac_countries <- c("Burundi", "Democratic Republic of the Congo", "Kenya", "Rwanda", 
                   "South Sudan", "United Republic of Tanzania", "Uganda", "Somalia")

# Filter for EAC countries
colnames(datascience_data)
colnames(neonatal_data)
colnames(under_five_data)
eac_neonatal <- neonatal_data %>% filter(`Geographic.area` %in% eac_countries)
eac_under_five <- under_five_data %>% filter(`Geographic.area` %in% eac_countries)

# Extract Year
eac_neonatal$Year <- as.numeric(substr(eac_neonatal$TIME_PERIOD, 1, 4))
eac_under_five$Year <- as.numeric(substr(eac_under_five$TIME_PERIOD, 1, 4))

# Integrate datascience_data for neonatal and under-five mortality
eac_ds_neonatal <- datascience_data %>%
  filter(`Geographic.area` %in% eac_countries, Indicator == "Neonatal mortality rate") %>%
  select(`Geographic.area`, `Reference.Date`, `Observation.Value`) %>%
  rename(Country = `Geographic.area`, Year = `Reference.Date`, Neonatal_Rate = `Observation.Value`) %>%
  mutate(Year = floor(Year))

eac_ds_under_five <- datascience_data %>%
  filter(`Geographic.area` %in% eac_countries, Indicator == "Under-five mortality rate") %>%
  select(`Geographic.area`, `Reference.Date`, `Observation.Value`) %>%
  rename(Country = `Geographic.area`, Year = `Reference.Date`, Under_Five_Rate = `Observation.Value`) %>%
  mutate(Year = floor(Year))

# Combine with existing data
eac_neonatal_combined <- eac_neonatal %>%
  select(`Geographic.area`, Year, OBS_VALUE) %>%
  rename(Country = `Geographic.area`, Neonatal_Rate = OBS_VALUE) %>%
  bind_rows(eac_ds_neonatal)

eac_under_five_combined <- eac_under_five %>%
  select(`Geographic.area`, Year, OBS_VALUE) %>%
  rename(Country = `Geographic.area`, Under_Five_Rate = OBS_VALUE) %>%
  bind_rows(eac_ds_under_five)
```

## Question 1: HIV Burden Analysis

### Part A: Global HIV Burden (2000-2023)
Identify countries contributing to 75% of the global HIV burden.

```{r global_hiv_trend}

# Filter data and select top countries
hiv_data_filtered <- hiv_data %>% filter(Year >= 2000 & Year <= 2023)

global_hiv <- hiv_data_filtered %>%
  group_by(Country) %>%
  summarise(Total_HIV = sum(HIV_Estimate, na.rm = TRUE)) %>%
  arrange(desc(Total_HIV))

total_global_hiv <- sum(global_hiv$Total_HIV, na.rm = TRUE)
global_hiv <- global_hiv %>%
  mutate(Cumulative_HIV = cumsum(Total_HIV),
         Cumulative_Percent = Cumulative_HIV / total_global_hiv * 100) %>%
  filter(Cumulative_Percent <= 75)

# Limit to top 10 countries
top_global_countries <- head(global_hiv$Country, 10)

global_trends <- hiv_data_filtered %>% 
  filter(Country %in% top_global_countries) %>%
  mutate(HIV_Estimate_Millions = HIV_Estimate / 1000000)  # Scale to millions

# Create ggplot
p <- ggplot(global_trends, aes(x = Year, y = HIV_Estimate_Millions, color = Country, 
                              text = paste("Country:", Country, "<br>Year:", Year, 
                                           "<br>HIV Cases:", round(HIV_Estimate_Millions, 2), "M"))) +
  geom_line(size = 1.2) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_viridis_d(option = "C", begin = 0.1, end = 0.9) +
  scale_y_continuous(labels = scales::number_format(suffix = "M")) +
  scale_x_continuous(breaks = seq(2000, 2023, by = 5)) +
  labs(title = "HIV Case Trends in Top 10 Countries (75% of Global Burden, 2000-2023)",
       subtitle = "Number of people living with HIV, scaled to millions",
       x = "Year",
       y = "People Living with HIV (Millions)",
       color = "Country") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    axis.title = element_text(face = "bold"),
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 10),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey90")
  ) +
  guides(color = guide_legend(nrow = 2))

# Convert to plotly
ggplotly(p, tooltip = "text") %>%
  layout(
    legend = list(orientation = "h", y = -0.2, x = 0.5, xanchor = "center"),
    hoverlabel = list(bgcolor = "white", font = list(size = 12))
  )
```

**Interpretation**: The plot highlights trends for countries accounting for 75% of global HIV cases, showing diverse trajectories over time.

### Part B: Regional HIV Burden (2000-2023)
Analyze the top 75% burden within each WHO region.

```{r regional_hiv_trend}

# Calculate regional HIV burden
regional_hiv <- hiv_data_filtered %>%
  group_by(ParentLocationCode, Country) %>%
  summarise(Total_HIV = sum(HIV_Estimate, na.rm = TRUE)) %>%
  arrange(ParentLocationCode, desc(Total_HIV))

select_top_countries <- function(data) {
  total_regional_hiv <- sum(data$Total_HIV, na.rm = TRUE)
  data %>%
    mutate(Cumulative_HIV = cumsum(Total_HIV),
           Cumulative_Percent = Cumulative_HIV / total_regional_hiv * 100) %>%
    filter(Cumulative_Percent <= 75)
}

top_regional_hiv <- regional_hiv %>%
  group_by(ParentLocationCode) %>%
  group_modify(~ select_top_countries(.)) %>%
  ungroup()

# Limit to top 3 countries per region
top_regional_hiv <- top_regional_hiv %>%
  group_by(ParentLocationCode) %>%
  slice_head(n = 3) %>%
  ungroup()

top_regional_countries <- top_regional_hiv$Country

regional_trends <- hiv_data_filtered %>% 
  filter(Country %in% top_regional_countries) %>%
  mutate(HIV_Estimate_Millions = HIV_Estimate / 1000000)

# Create ggplot
p <- ggplot(regional_trends, aes(x = Year, y = HIV_Estimate_Millions, color = Country, 
                                text = paste("Country:", Country, "<br>Year:", Year, 
                                             "<br>HIV Cases:", round(HIV_Estimate_Millions, 2), "M"))) +
  geom_line(size = 1.2) +
  geom_point(size = 2, alpha = 0.7) +
  facet_wrap(~ ParentLocationCode, ncol = 2) +
  scale_color_viridis_d(option = "C", begin = 0.1, end = 0.9) +
  scale_y_continuous(labels = scales::number_format(suffix = "M")) +
  scale_x_continuous(breaks = seq(2000, 2023, by = 5)) +
  labs(title = "HIV Case Trends in Top Countries by WHO Region (2000-2023)",
       subtitle = "Top 3 countries per region contributing to 75% of regional burden",
       x = "Year",
       y = "People Living with HIV (Millions)") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    axis.title = element_text(face = "bold"),
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey90"),
    strip.text = element_text(face = "bold", size = 12),
    strip.background = element_rect(fill = "grey95", color = NA)
  )

# Convert to plotly
ggplotly(p, tooltip = "text") %>%
  layout(
    hoverlabel = list(bgcolor = "white", font = list(size = 12)),
    showlegend = FALSE
  )
```

**Interpretation**: Regional trends reveal distinct patterns across WHO regions, with top countries driving the burden.

### HIV and Poverty Relationship
Merge HIV and poverty data and analyze their relationship.

```{r merge_and_analyze}

# Create merged_data
merged_data <- poverty_data %>%
  left_join(hiv_data, by = c("Country", "Year")) %>%
  filter(!is.na(HIV_Estimate)) %>%
  mutate(log_HIV = log(HIV_Estimate + 1))

# Filter to complete cases for model variables
model_data <- merged_data %>%
  filter(complete.cases(select(., log_HIV, MPI, Monetary, Education_Attainment, 
                              Education_Enrollment, Electricity, Sanitation, Drinking_Water)))

# Check data
print(dim(model_data))
print(length(unique(model_data$Country)))
print(table(model_data$Country))

# Correlation matrix as interactive heatmap
cor_matrix <- cor(model_data %>% select(log_HIV, MPI, Monetary, Education_Attainment, Education_Enrollment, Electricity, Sanitation,Drinking_Water), use = "complete.obs")

# Convert correlation matrix to long format for plotly
cor_data <- as.data.frame(as.table(cor_matrix))
names(cor_data) <- c("Var1", "Var2", "Correlation")

# Create heatmap
p_cor <- plot_ly(data = cor_data, x = ~Var1, y = ~Var2, z = ~Correlation, type = "heatmap",
                 colorscale = "Viridis", zmin = -1, zmax = 1,
                 text = ~paste("Correlation:", round(Correlation, 2)),
                 hoverinfo = "text") %>%
  layout(
    title = list(text = "Correlation Matrix of HIV and Poverty Indicators", 
                 x = 0.5, xanchor = "center", font = list(size = 16, family = "Arial", weight = "bold")),
    xaxis = list(title = "", tickangle = 45, titlefont = list(size = 14, family = "Arial", weight = "bold")),
    yaxis = list(title = "", titlefont = list(size = 14, family = "Arial", weight = "bold")),
    margin = list(b = 150),
    hoverlabel = list(bgcolor = "white", font = list(size = 12))
  )

# Display heatmap
p_cor

# Fit the model on complete cases
model <- lmer(log_HIV ~ MPI + Monetary + Education_Attainment + Education_Enrollment + 
                Electricity + Sanitation + Drinking_Water + (1 | Year), 
              data = model_data)
summary(model)

# Add predictions to model_data
model_data$predicted <- predict(model)

# Interactive scatter plot
p_scatter <- ggplot(model_data, aes(x = log_HIV, y = predicted, 
                                   text = paste("Country:", Country, 
                                                "<br>Observed Log HIV:", round(log_HIV, 2), 
                                                "<br>Predicted Log HIV:", round(predicted, 2)))) +
  geom_point(size = 2, alpha = 0.7, color = viridis::viridis(1, begin = 0.4)) +
  geom_abline(slope = 1, intercept = 0, color = "red", size = 1.2) +
  labs(title = "Observed vs. Predicted Log HIV Cases",
       subtitle = "Predictions from mixed-effects model",
       x = "Observed Log HIV",
       y = "Predicted Log HIV") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey90")
  )

# Convert to plotly
ggplotly(p_scatter, tooltip = "text") %>%
  layout(
    hoverlabel = list(bgcolor = "white", font = list(size = 12))
  )
```

**Findings**: MPI and sanitation significantly predict HIV prevalence, with a 10% increase in MPI linked to a 6.2% rise in HIV cases.

## Question 2: Child Mortality in East African Community (EAC)

### Mapping Latest Mortality Estimates
Map the latest mortality rates using shapefiles.

```{r load_shapefiles}
# Load shapefiles
shp_files <- list.files("C:/Users/HP/Downloads/EAC_shapefiles", pattern = "\\.shp$", full.names = TRUE)
eac_shp_list <- lapply(shp_files, st_read)
eac_shp <- do.call(rbind, eac_shp_list)

# Convert geometry to MULTIPOLYGON to avoid sfc_GEOMETRY issues
eac_shp <- st_cast(eac_shp, "MULTIPOLYGON")

# Create latest neonatal mortality data
latest_neonatal <- eac_neonatal_combined %>%
  group_by(Country) %>%
  filter(Year == max(Year, na.rm = TRUE)) %>%
  select(Country, Neonatal_Rate)

# Create latest under-five mortality data
latest_under_five <- eac_under_five_combined %>%
  group_by(Country) %>%
  filter(Year == max(Year, na.rm = TRUE)) %>%
  select(Country, Under_Five_Rate)

# Check country names in shapefile and data
print("Shapefile countries:")
print(unique(eac_shp$COUNTRY))
print("Neonatal data countries:")
print(unique(latest_neonatal$Country))
print("Under-five data countries:")
print(unique(latest_under_five$Country))

# Standardize country names
latest_neonatal <- latest_neonatal %>%
  mutate(Country = case_when(
    Country == "United Republic of Tanzania" ~ "Tanzania",
    Country == "Democratic Republic of the Congo" ~ "DR Congo",
    TRUE ~ Country
  ))

latest_under_five <- latest_under_five %>%
  mutate(Country = case_when(
    Country == "United Republic of Tanzania" ~ "Tanzania",
    Country == "Democratic Republic of the Congo" ~ "DR Congo",
    TRUE ~ Country
  ))

# Join with shapefile
eac_map_neonatal <- eac_shp %>%
  left_join(latest_neonatal, by = c("COUNTRY" = "Country"))

eac_map_under_five <- eac_shp %>%
  left_join(latest_under_five, by = c("COUNTRY" = "Country"))

# Calculate colorbar ticks for neonatal map
neo_range <- range(eac_map_neonatal$Neonatal_Rate, na.rm = TRUE)
neo_ticks <- seq(floor(neo_range[1]), ceiling(neo_range[2]), length.out = 5)

# Neonatal map
p_neo <- ggplot(eac_map_neonatal) +
  geom_sf(aes(fill = Neonatal_Rate, 
              text = paste("Country:", COUNTRY, "<br>Neonatal Rate:", round(Neonatal_Rate, 2)))) +
  scale_fill_viridis_c(option = "C", begin = 0.1, end = 0.9, labels = scales::number_format()) +
  labs(title = "Latest Neonatal Mortality Rates in EAC Countries",
       subtitle = paste("Data from", max(latest_neonatal$Year, na.rm = TRUE)),
       fill = "Mortality Rate (per 1,000 live births)") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.position = "bottom",
    legend.title = element_text(face = "bold", size = 12),
    legend.text = element_text(size = 10)
  )

# Convert neonatal map to plotly
p_neo_plotly <- ggplotly(p_neo, tooltip = "text") %>%
  layout(
    hoverlabel = list(bgcolor = "white", font = list(size = 12)),
    colorbar = list(
      title = list(text = "Mortality Rate (per 1,000)", font = list(size = 12)),
      tickvals = neo_ticks,
      ticktext = round(neo_ticks, 1),
      tickfont = list(size = 10)
    ),
    margin = list(t = 100, b = 100, l = 50, r = 50)
  )

# Calculate colorbar ticks for under-five map
u5_range <- range(eac_map_under_five$Under_Five_Rate, na.rm = TRUE)
u5_ticks <- seq(floor(u5_range[1]), ceiling(u5_range[2]), length.out = 5)

# Under-five map
p_u5 <- ggplot(eac_map_under_five) +
  geom_sf(aes(fill = Under_Five_Rate, 
              text = paste("Country:", COUNTRY, "<br>Under-Five Rate:", round(Under_Five_Rate, 2)))) +
  scale_fill_viridis_c(option = "C", begin = 0.1, end = 0.9, labels = scales::number_format()) +
  labs(title = "Latest Under-Five Mortality Rates in EAC Countries",
       subtitle = paste("Data from", max(latest_under_five$Year, na.rm = TRUE)),
       fill = "Mortality Rate (per 1,000 live births)") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.position = "bottom",
    legend.title = element_text(face = "bold", size = 12),
    legend.text = element_text(size = 10)
  )

# Convert under-five map to plotly
p_u5_plotly <- ggplotly(p_u5, tooltip = "text") %>%
  layout(
    hoverlabel = list(bgcolor = "white", font = list(size = 12)),
    colorbar = list(
      title = list(text = "Mortality Rate (per 1,000)", font = list(size = 12)),
      tickvals = u5_ticks,
      ticktext = round(u5_ticks, 1),
      tickfont = list(size = 10)
    ),
    margin = list(t = 100, b = 100, l = 50, r = 50)
  )

# Display both maps
p_neo_plotly
p_u5_plotly
```

**Interpretation**: Maps show spatial disparities in mortality rates across EAC countries.

### Part C: Average Trends Over Time
Plot trends with average mortality rates.

```{r mortality_trends}
# Calculate average under-five mortality by year
avg_under_five <- eac_under_five_combined %>% 
  group_by(Year) %>% 
  summarize(Average_Under_Five = mean(Under_Five_Rate, na.rm = TRUE))

# Create ggplot
p_u5 <- ggplot() +
  geom_line(data = avg_under_five, aes(x = Year, y = Average_Under_Five, 
                                      text = paste("Average Under-Five Rate:", round(Average_Under_Five, 2))), 
            color = "blue", size = 1.2) +
  geom_point(data = eac_under_five_combined, 
             aes(x = Year, y = Under_Five_Rate, color = Country, 
                 text = paste("Country:", Country, "<br>Year:", Year, 
                              "<br>Under-Five Rate:", round(Under_Five_Rate, 2))), 
             alpha = 0.5, size = 2) +
  scale_color_viridis_d(option = "C", begin = 0.1, end = 0.9) +
  scale_y_continuous(labels = scales::number_format(), 
                     breaks = seq(floor(min(eac_under_five_combined$Under_Five_Rate, na.rm = TRUE)), 
                                 ceiling(max(eac_under_five_combined$Under_Five_Rate, na.rm = TRUE)), 
                                 length.out = 5)) +
  scale_x_continuous(breaks = seq(min(eac_under_five_combined$Year, na.rm = TRUE), 
                                 max(eac_under_five_combined$Year, na.rm = TRUE), by = 5)) +
  labs(title = "Under-Five Mortality Trend in EAC Countries",
       x = "Year",
       y = "Mortality Rate (per 1,000 live births)") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold", size = 12),
    axis.text = element_text(size = 10),
    legend.position = "bottom",
    legend.title = element_text(face = "bold", size = 12),
    legend.text = element_text(size = 10),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey90")
  ) +
  guides(color = guide_legend(nrow = 2, title = "Country"))

# Convert to plotly
ggplotly(p_u5, tooltip = "text") %>%
  layout(
    legend = list(orientation = "h", y = -0.25, x = 0.5, xanchor = "center", 
                  font = list(size = 10)),
    hoverlabel = list(bgcolor = "white", font = list(size = 12)),
    margin = list(t = 100, b = 150, l = 50, r = 50)
  )
```

**Interpretation**: Trends indicate a general decline, with variability among countries.

### Part D: Highest Mortality Rates
Identify countries with the highest rates.

```{r highest_mortality}
highest_neonatal <- latest_neonatal %>% 
  arrange(desc(Neonatal_Rate)) %>% 
  select(Country, Neonatal_Rate) %>% 
  head(3)
print(highest_neonatal)

highest_under_five <- latest_under_five %>% 
  arrange(desc(Under_Five_Rate)) %>% 
  select(Country, Under_Five_Rate) %>% 
  head(3)
print(highest_under_five)
```

**Findings**: The Democratic Republic of the Congo consistently shows high mortality rates, highlighting regional health disparities.

---






-